My wonderful paper

### 2024-12-04 _manuscript/_tex/index.tex
literature review on data analysis checks

\subsection{Data analysis checks in statistical
software}\label{data-analysis-checks-in-statistical-software}

Currently, little has been done on how to computationally incorporate
this diagnostic approach into data analysis workflow or software. Most
of the diagnostic tools focus on defining user-specified rules, such as
data quality checks or producing metrics to summarize model performance,
as in model diagnostics. For example, the \texttt{assertr} package
\citep{assertr} and the \texttt{pointblank} package \citep{pointblank}
provide data validation tools to allow users to define data quality
checks. In contrast, packages provide model checks tools like
\texttt{performance} \citep{performance} and \texttt{yardstick}
\citep{yardstick}, from the \texttt{tidymodels} \citep{tidymodels}
ecosystem, offers goodness-of-fit, residual analysis, and model
performance metrics.

These packages provide the tools to conduct diagnostics but still don't
reflect the mental process of how data analysts diagnose the unexpected
output. For example, when an unexpected output occurs, an analyst may
check on whether a column in the data frame is between two values for
data quality. However, it is not documented what motivates the analyst
to conduct this check -- whether it also applies to other researchers
analyzing new data in the same contexts, whether it is a common practice
in the field, or whether it is a reaction to this particular data or
scenario. Currently, most of these assumptions are not discussed in the
publication or captured by tools themselves. While one might be able to
infer some of the mental process of the analysts from external sources,
such as talking to them or watching screencast videos produced by the
analysts e.g.~TidyTuesday screencast videos, these insights are not
systematically documented or made machine-readable. This gap highlights
the need for tools that provide higher level documentation of reasoning
behind the checks, facilitating a more transparent and interpretable
analysis process.

>>>>

\subsection{Data analysis checks}\label{data-analysis-checks}

A substantial body of literature has addressed the definition of data
quality \citep[more]{8642813} and developed frameworks, that includes
dimensions, attributes, and measures to evaluate and improve data
quality
\citep{cai2015challenges, wang1996beyond, 6204995, woodall2014classification}.
These frameworks are often used information system and database
management and support business decision-making in the industry. For
research purposes, high-quality data ensures the credibility of
scientific findings and supports reproducibility and reusability in
future studies {[}ref{]}. With the growing prevalence of open data in
scientific research, the consumers or users of the data typically are no
longer the data producers or collectors who have the full knowledge of
data in hand, prompting more interest towards the data quality checks in
the data analysis process. In R, there are some packages,
like\texttt{skimr} \citep{skimr} and \texttt{dataMaid} \citep{dataMaid},
provides basic data screening and reporting, while another class of
packages, e.g.~\texttt{assertr} \citep{assertr}, \texttt{validate}
\citep{validate}, and \texttt{pointblank} \citep{pointblank}, focuses on
providing data validation tools, allowing users to define customized
data quality checks based on the applications.

While data validation focuses on verifying or subsetting data based on
predefined rules \citep{zio2015methodology}, data analysis checks
specify assumptions made on the data required for the analysis process
to obtain the reported results. These checks often involve evaluating
aspects such as data distribution and outliers, relationships between
variables, and the validity of model assumptions.

However, there are other checks that reflect the mental process of of
how data analysts diagnose the unexpected output that are not documented
in this process. For example, when an unexpected output occurs, an
analyst may check on whether a column in the data frame is between two
values for data quality. However, it is not documented what motivates
the analyst to conduct this check -- whether it also applies to other
researchers analyzing new data in the same contexts, whether it is a
common practice in the field, or whether it is a reaction to this
particular data or scenario. Currently, most of these assumptions are
not discussed in the publication or captured by tools themselves. While
one might be able to infer some of the mental process of the analysts
from external sources, such as talking to them or watching screencast
videos produced by the analysts e.g.~TidyTuesday screencast videos,
these insights are not systematically documented or made
machine-readable. A question to consider here is whether it is possible
to deliver this type of quality checks as structured data in a data
analysis. This gap highlights the need for tools that provide higher
level documentation of reasoning behind the checks, facilitating a more
transparent and interpretable analysis process.

### 2024-12-04 index.qmd
literature review on data analysis checks

## Data analysis checks in statistical software

Currently, little has been done on how to computationally incorporate this diagnostic approach into data analysis workflow or software. Most of the diagnostic tools focus on defining user-specified rules, such as data quality checks or producing metrics to summarize model performance, as in model diagnostics. For example, the `assertr` package [@assertr] and the `pointblank` package [@pointblank] provide data validation tools to allow users to define data quality checks. In contrast, packages provide model checks tools like `performance` [@performance] and `yardstick` [@yardstick], from the `tidymodels` [@tidymodels] ecosystem, offers goodness-of-fit, residual analysis, and model performance metrics.

These packages provide the tools to conduct diagnostics but still don't reflect the mental process of how data analysts diagnose the unexpected output. For example, when an unexpected output occurs, an analyst may check on whether a column in the data frame is between two values for data quality. However, it is not documented what motivates the analyst to conduct this check -- whether it also applies to other researchers analyzing new data in the same contexts, whether it is a common practice in the field, or whether it is a reaction to this particular data or scenario. Currently, most of these assumptions are not discussed in the publication or captured by tools themselves. While one might be able to infer some of the mental process of the analysts from external sources, such as talking to them or watching screencast videos produced by the analysts e.g. TidyTuesday screencast videos, these insights are not systematically documented or made machine-readable. This gap highlights the need for tools that provide higher level documentation of reasoning behind the checks, facilitating a more transparent and interpretable analysis process.

>>>>

## Data analysis checks

A substantial body of literature has addressed the definition of data quality [@8642813, more] and developed frameworks, that includes dimensions, attributes, and measures to evaluate and improve data quality [@cai2015challenges; @wang1996beyond; @6204995; @woodall2014classification]. These frameworks are often used information system and database management and support business decision-making in the industry. For research purposes, high-quality data ensures the credibility of scientific findings and supports reproducibility and reusability in future studies [ref]. With the growing prevalence of open data in scientific research, the consumers or users of the data typically are no longer the data producers or collectors who have the full knowledge of data in hand, prompting more interest towards the data quality checks in the data analysis process. In R, there are some packages, like`skimr` [@skimr] and `dataMaid` [@dataMaid], provides basic data screening and reporting, while another class of packages, e.g. `assertr` [@assertr], `validate` [@validate], and `pointblank` [@pointblank], focuses on providing data validation tools, allowing users to define customized data quality checks based on the applications.

While data validation focuses on verifying or subsetting data based on predefined rules [@zio2015methodology], data analysis checks specify assumptions made on the data required for the analysis process to obtain the reported results. These checks often involve evaluating aspects such as data distribution and outliers, relationships between variables, and the validity of model assumptions.

However, there are other checks that reflect the mental process of of how data analysts diagnose the unexpected output that are not documented in this process. For example, when an unexpected output occurs, an analyst may check on whether a column in the data frame is between two values for data quality. However, it is not documented what motivates the analyst to conduct this check -- whether it also applies to other researchers analyzing new data in the same contexts, whether it is a common practice in the field, or whether it is a reaction to this particular data or scenario. Currently, most of these assumptions are not discussed in the publication or captured by tools themselves. While one might be able to infer some of the mental process of the analysts from external sources, such as talking to them or watching screencast videos produced by the analysts e.g. TidyTuesday screencast videos, these insights are not systematically documented or made machine-readable. A question to consider here is whether it is possible to deliver this type of quality checks as structured data in a data analysis. This gap highlights the need for tools that provide higher level documentation of reasoning behind the checks, facilitating a more transparent and interpretable analysis process.
